Quantitative Benchmarking for Automatically Generated Conjectures

### 2017-11-16 TEBenchmarkPaper/article.tex
More changes

The precision and recall for multiple samples $S$ can be combined in two ways:
finding their mean value (the \emph{average of the ratios}) or by summing the
numerators and denominators (the \emph{ratio of the averages}). The former gives
the expected precision and recall for a run, which is of direct relevance to
users of these tools, and is hence our preferred measure. The latter assigns more weight to theories which generate more
conjectures, but that is not justification there is no reason to consider these theories ``better'',

@@

Another advantage to our statistical approach is that we can compute the
\emph{variance} of our data, which is unknown in the one-off evaluations which
have been performed previously. In particular, we might be concerned that such
previous evaluations used ``popular'' theories, like those involving natural
numbers, which may be unrepresentative of typical usage: any number theorist
knows how amenable natural numbers are to conjecture and exploration, and we
should not expect a typical user's theory (e.g. a software library) to be
similarly rich.

@@

We have applied our methodology to evaluating QuickSpec, %TODO:cite
a library for conjecturing equations about functions in the Haskell programming
language. We use the Tons of Inductive Problems (TIP) theorem proving benchmark
to generate our corpus.

@@

We use QuickSpec version 0.9.6 to demonstrate our benchmarking methodology. This
is a library for the Haskell programming language, which is given a
\emph{signature} containing Haskell functions and (universally quantified) typed
variables, and outputs conjectured equations involving these terms.

The QuickSpec algorithm first enumerates all well-typed expressions, up to a
certain depth (3 by default), involving the given terms. Expressions of the same
type are assumed to be equivalent, and grouped into equivalence classes. A
testing process then tries to disprove this assumption by instantiating each
variable with a randomly chosen value (using the QuickCheck testing library),
evaluating the expressions and comparing the results. The equivalence classes
are split to separate any expressions which are observed to be different, and
the process repeats with new random values.

@@

\begin{itemize}
\item We must decide what variables to include. We add three for each type that
  appears as a function argument, except for types which have no QuickCheck data
  generator.
\item We must \emph{monomorphise} all types. For example, functions like
  \texttt{cons} are \emph{polymorphic}: they build lists of any element type.
  We need to pick a specific type in order to generate random values, so we
  resolve this, arbitrarily, by picking
  \emph{Integer}.~\footnote{We pick \texttt{Integer} for variables of kind
    \texttt{*}, and \texttt{[]} (Haskell's list type constructor) for those of
    kind \texttt{* -> *}. If these wouldn't satisfy some constraint, we pick a
    suitable type non-deterministically from the module scope, or skip these
    variables completely if none is found.}
\item Haskell functions only take one argument, as as in the $\lambda$-calculus,
  so functions of multiple arguments must be simulated by currying.
  Unfortunately QuickSpec cannot compare functions for equality, so we must give
  each function in our signature an explicit arity. Arities from zero to five
  are supported, so we avoid partial application by picking the highest that is
  type-correct.
\end{itemize}

@@

We use TIP version 0.2 which contains 343 problems, each stating a single
theorem and together defining a total of 618 datatypes and 1498 functions. Most
of these are duplicates, since each problem (re-)defines all of the datatypes
and functions it involves. We prefix the names of all definitions with the
filename it came from, and collect them into one big set.

Datatypes can have several ``constructors'' and ``destructors''. For example the
natural numbers can be defined as follows:

@@

Since our GHC plugin only lists function definitions, we add separate function
definitions for each constructor and destructor (these are just
$\eta$-expansions).

@@

We applied our sampling methodology to the 269 TIP definitions for sample sizes
between 1 and 20, above which point QuickSpec becomes very slow. We chose 30
samples for each size, but the smaller sizes included repetitions due to the
requirement for samples to admit at least one theorem. The results are shown in
figure~\ref{figure:quickspec_results}.

@@

Theorem statements from TIP were first updated to reference only the 269
distinct names we explored, and references to constructors and destructors were
updated to use the $\eta$-expanded versions instead. Equations were normalised
in the same way as above (non-equations were kept unchanged), and elements of
these two sets were deemed to match if they were syntactically identical.

We have applied our benchmarking methodology to the Tons of Inductive Problems
(TIP) problem set for automated theorem provers. TIP satisfies the desiderata of
section\ref{section:prep} since each problem has standalone type and function
definitions, making their separation trivial. The problem set includes known
examples from the software verification and inductive theorem proving
literature, including higher-order functions, common datatypes, etc., ensuring
that benchmark results are relevant to those fields. It also includes tools to
convert the problems into a variety of languages, including Haskell and
Isabelle.

TIP version 0.2 contains FIXME theorem statements, each of which is accompanied
by definitions of the types and functions it uses (other than those for integers
and booleans, which are ``built in''). There are FIXME function definitions, of
which most are duplicates. We remove $\alpha$-equivalent definitions in the
following way:

\begin{enumerate}
\item We prefix all type and function names with the filename in which we found
  the definition, so that e.g. the theorem statement from
  \texttt{tip2015/int\_add\_assoc.smt2} references the \texttt{plus} function from
  that file, rather than e.g. the \texttt{plus} function from
  \texttt{isaplanner/prop\_07.smt2}.
\item For each definition, all local variables (including function arguments,
  constructor parameters, pattern-matching variables and \texttt{let} bindings)
  are replaced by standard names \texttt{local-1}, \texttt{local-2}, etc. in a
  well-defined order.
\item An $O(n^2)$ comparison of each definition to every other definition is
  performed, to find any which are syntactically-identical modulo the names
  being defined (but \emph{not} the names being referenced, i.e. ). For those which are,
  we record which names are equivalent, replace all references to these names
  with whichever is smallest lexicographically, and remove
\end{enumerate}

Altogether TIP problem by QuickSpec and IsaCoSy theory
exploration systems. To assess the scalability of their algorithms, we sampled
theories of various sizes: from theories containing only a single definition, up
to theories containing 20 definitions.
 We call the Theory Exploration Benchmark (TEB).

To do this we can take a set of ATP tasks, for example proving the equation
\ref{eq:mapreduce} from the theory in figure \ref{figure:list_theory}, and we
split apart the definitions (to form a theory) from the theorem statements (to
form our ground truth).

We focus on the Tons of Inductive Problems (TIP) theorem proving
benchmark~\cite{claessen2015tip}, since it has several desirable properties:

All together, TIP provides 219 distinct function definitions and 343
  theorem statements, which is enough to fully exercise current ATE systems.
Benchmark problems include examples from the theorem proving literature
  as well as common program verification tasks, ensuring the resulting corpus is
  relevant to researchers and practitioners.
TIP's maintainers provide a set of tools, including translators from TIP's
  native format (based on SMT-Lib~\cite{BarFT-SMTLIB}) to Haskell and Isabelle
  (which are used by ATE systems).

To aid reproducibility, we use a cryptographic hash as our source of randomness:
the $n$th sample of size $s$ uses the SHA256 hash of $n$, $s$ and the definition
corpus; the latter prevents tailoring the corpus to influence the choices,
although ATE systems could .

To provide a more robust comparison of these two systems, we also applied a
paired difference test: measuring, for each sampled theory, the difference in
time, precision and recall between the two systems.

@@

\begin{acknowledgements}
  We are grateful to Jianguo Zhang for help with our statistical analysis.

  EPSRC
\end{acknowledgements}

>>>>

The precision and recall for multiple samples $S$ can be combined in two ways:
finding their mean value (the \emph{average of the ratios}) or by summing the
numerators and denominators (the \emph{ratio of the averages}). The former gives
the expected precision and recall for a run, which is of direct relevance to
users of these tools, and is hence our preferred measure. The latter assigns
more weight to theories which generate more conjectures, but we avoid this since
there is no \emph{a priori} justification for considering such theories as
``better''.

@@

Another advantage to our statistical approach is that we can compute the
\emph{variance} of our data, which is unknown in the previous one-off
evaluations. In particular, we might be concerned that such previous evaluations
used ``popular'' theories, like those involving natural numbers, which may be
unrepresentative of typical usage: we should not expect a typical user's theory
(e.g. a software library) to be as mathematically rich as number theory!

@@

We have applied our methodology to evaluating QuickSpec, %TODO:cite
a library for conjecturing equations about functions written in the Haskell
programming language. We use the Tons of Inductive Problems (TIP) theorem
proving benchmark to generate our corpus.

@@

We use QuickSpec version 0.9.6 to demonstrate our benchmarking methodology. This
is a Haskell library which takes a \emph{signature} containing Haskell functions
and (universally quantified) typed variables, and outputs conjectured equations
involving these terms.

The QuickSpec algorithm first enumerates all well-typed expressions up to a
certain depth (3 by default), involving the given terms. Expressions of the same
type are assumed to be equivalent, and grouped into equivalence classes. A
testing process then tries to disprove this assumption by instantiating each
variable with a randomly chosen value (using the QuickCheck testing library),
evaluating the expressions and comparing the results. The equivalence classes
are split to separate any expressions which are observed to be different, and
the process repeats with new random values.

@@

\begin{itemize}
\item We must decide what variables to include. We add three for each type that
  appears as a function argument, except for types which have no QuickCheck data
  generator.
\item We must \emph{monomorphise} all types. For example, functions like
  \texttt{cons} are \emph{polymorphic}: they build lists of any element type.
  We need to pick a specific type in order to generate random values, so we
  resolve this (arbitrarily) by picking
  \texttt{Integer}.~\footnote{We pick \texttt{Integer} for variables of kind
    \texttt{*}, and \texttt{[]} (Haskell's list type constructor) for those of
    kind \texttt{* -> *}. If these violate some constraint, we pick a suitable
    type non-deterministically from the module scope, or skip these variables
    completely if none is found.}
\item Haskell functions only take one argument, as as in the $\lambda$-calculus,
  so functions of multiple arguments are simulated by currying.
  Unfortunately QuickSpec cannot compare functions for equality, so we must give
  each function in our signature an explicit arity. Arities from zero to five
  are supported, so we avoid partial application by picking the highest that is
  type-correct.
\end{itemize}

@@

We chose TIP for our ground truth since it satisfies the desiderata of
section~\ref{section:prep}: each problem has standalone type and function
definitions, making their separation trivial; known examples from the software
verification and inductive theorem proving literature are included, ensuring
relevance to those fields; the format includes higher-order functions and
inductive datatypes, which we are interested in; plus it is accompanied by
tooling to convert this format into a variety of languages, including Haskell
and Isabelle.

We use TIP version 0.2 which contains 343 problems, each stating a single
theorem and together defining a total of 618 datatypes and 1498 functions. Most
of these are duplicates, since each problem (re-)defines all of the datatypes
and functions it involves. To avoid ambiguity we prefix the names of all
definitions with the filename it came from, and collect them into one big set.

TIP datatypes can have several ``constructors'' and ``destructors''. For example
the natural numbers can be defined as follows:

@@

Since our GHC plugin doesn't list constructors or destructors, we add separate
function definitions for each (these are just $\eta$-expansions).

@@

We applied our sampling methodology to the 269 TIP definitions for sample sizes
between 1 and 20, above which point QuickSpec becomes very slow. We chose 30
samples for each size, but the smaller sizes included repetitions due to the
requirement for samples to admit at least one theorem. A timeout of 3 minutes
was imposed, since preliminary experiments showed that most runs either finish
before this time, or still do not finish after an hour. The results are shown in
figure~\ref{figure:quickspec_results}.

@@

Theorem statements from TIP were first updated to reference only the 269
distinct names we explored, and references to constructors and destructors were
updated to use the $\eta$-expanded versions instead. Equations were normalised
in the same way as above, whilst non-equations and those containing
\textt{lambda} functions (which QuickSpec cannot synthesise) were replaced with
null values. Elements of these two sets were deemed to match if they were
syntactically identical.

% Altogether TIP problem by QuickSpec and IsaCoSy theory
% exploration systems. To assess the scalability of their algorithms, we sampled
% theories of various sizes: from theories containing only a single definition, up
% to theories containing 20 definitions.
%  We call the Theory Exploration Benchmark (TEB).

% To do this we can take a set of ATP tasks, for example proving the equation
% \ref{eq:mapreduce} from the theory in figure \ref{figure:list_theory}, and we
% split apart the definitions (to form a theory) from the theorem statements (to
% form our ground truth).

% We focus on the Tons of Inductive Problems (TIP) theorem proving
% benchmark~\cite{claessen2015tip}, since it has several desirable properties:

% All together, TIP provides 219 distinct function definitions and 343
%   theorem statements, which is enough to fully exercise current ATE systems.
% Benchmark problems include examples from the theorem proving literature
%   as well as common program verification tasks, ensuring the resulting corpus is
%   relevant to researchers and practitioners.
% TIP's maintainers provide a set of tools, including translators from TIP's
%   native format (based on SMT-Lib~\cite{BarFT-SMTLIB}) to Haskell and Isabelle
%   (which are used by ATE systems).

% To aid reproducibility, we use a cryptographic hash as our source of randomness:
% the $n$th sample of size $s$ uses the SHA256 hash of $n$, $s$ and the definition
% corpus; the latter prevents tailoring the corpus to influence the choices,
% although ATE systems could .

% To provide a more robust comparison of these two systems, we also applied a
% paired difference test: measuring, for each sampled theory, the difference in
% time, precision and recall between the two systems.

@@

%\begin{acknowledgements}
%  We are grateful to Jianguo Zhang for help with our statistical analysis.
%
%  EPSRC
%\end{acknowledgements}
